---
title: "Clean CB Replication"
output: 
---
Necessary packages: 
```{r}
library(tidyverse)

```

Building permit data from data.honolulu.gov 
```{r}
data_2 <- read_csv("data/Building_Permits_-_January_1__2005_through_March_31__2025_20250418 (1).csv")
```
Clean data
Fix dates. in createddate and issuedate. Dates currently in this format: 05/09/2023
```{r}
data_2 %>% 
  mutate(created_date = mdy(createddate)) %>% 
  select(created_date, createddate)

#make it permanent

data_2 <- data_2 %>% 
  mutate(created_date = mdy(createddate), .after = createddate)

#change dates for issuedate
data_2 %>% 
  mutate(issue_date = mdy(issuedate)) %>% 
  select(issue_date, issuedate)

#make it permanent 
data_2 <- data_2 %>% 
  mutate(issue_date = mdy(issuedate), .after = issue_date)
```

Get rid of dates prior to 2012 and after 2022.

```{r}
data_2 %>% 
  filter(created_date > 2012 & created_date < 2023) %>% 
  range(data_2$created_date)

#that did not work

data_2 %>% 
  filter(created_date >= as_date ("2012-01-01") & created_date <= as_date("2022-12-31")) %>%
  select(created_date) %>% 
  arrange(created_date)
#checking that the first date is Jan. 1, 2012

data_2 %>% 
  filter(created_date >= as_date ("2012-01-01") & created_date <= as_date("2022-12-31")) %>%
  select(created_date) %>% 
  arrange(desc(created_date))

#checking that the last date is December 31, 2022. 

#make it permanent 

data_2 <- data_2 %>% 
  filter(created_date >= as_date ("2012-01-01") & created_date <= as_date("2022-12-31"))
#I now have 235,944 rows in my data
        
```

Break data up into four categories representing some fo the most common types of project: 
1) New building with electrical, plumbing and solar 
2) Alteration with addition with electrical and plumbing work
3) Alteration with electrical and plumbing
4) Alteration with electrical work 

```{r}
#viewing column names
glimpse(data_2)
# relevant column names: 
#addition == ""
#alteration == ""
#plumbingwork == ""
#electricalwork == ""
#newbuilding == ""

#unclear if Civil Beat used the column "solarvpinstallation" or the column "solar". We know from earlier analyses these two columns are not always the same. 
```

Create new column called "permit_type" and categorize data with new labels. 
1) New building with electrical, plumbing and solar "NB_EPS" 
NB_EPS ~ newbuilding == "Y" & electricalwork == "Y" & plumbingwork == "Y" & "solar" == "Y" & addition == "N" & alteration == "N",

2) Alteration with addition with electrical and plumbing work
Alt_Add_EP ~ newbuilding == "N" & electricalwork == "Y" & plumbingwork == "Y" & "solar" == "N" & addition == "Y" & alteration == "Y", 

3) Alteration with electrical and plumbing
Alt_EP ~ newbuilding == "N" & electricalwork == "Y" & plumbingwork == "Y" & "solar" == "N" & addition == "N" & alteration == "Y",

4) Alteration with electrical work 
Alt_E ~ newbuilding == "N" & electricalwork == "Y" & plumbingwork == "N" & "solar" == "N" & addition == "N" & alteration == "Y",

```{r}
data_2 %>% 
  mutate(permit_type = case_when(
   newbuilding == "Y" & electricalwork == "Y" & plumbingwork == "Y" & solar == "Y" & addition == "N" & alteration == "N" ~ "NB_EPS",
   newbuilding == "N" & electricalwork == "Y" & plumbingwork == "Y" & solar == "N" & addition == "Y" & alteration == "Y" ~ "Alt_Add_EP", 
   newbuilding == "N" & electricalwork == "Y" & plumbingwork == "Y" & solar == "N" & addition == "N" & alteration == "Y" ~ "Alt_EP",
   newbuilding == "N" & electricalwork == "Y" & plumbingwork == "N" & solar == "N" & addition == "N" & alteration == "Y" ~ "Alt_E",
   .default = "Unknown"))
#I used .default = "Unknown" instead of TRUE ~ NA_character_ because we used .default = "Unknown" in class and TRUE ~ NA_character_ is ChatGPT's recommendation -- I don't trust chatGPT. 

data_2 %>% 
  mutate(permit_type = case_when(
   newbuilding == "Y" & electricalwork == "Y" & plumbingwork == "Y" & solar == "Y" & addition == "N" & alteration == "N" ~ "NB_EPS",
   newbuilding == "N" & electricalwork == "Y" & plumbingwork == "Y" & solar == "N" & addition == "Y" & alteration == "Y" ~ "Alt_Add_EP", 
   newbuilding == "N" & electricalwork == "Y" & plumbingwork == "Y" & solar == "N" & addition == "N" & alteration == "Y" ~ "Alt_EP",
   newbuilding == "N" & electricalwork == "Y" & plumbingwork == "N" & solar == "N" & addition == "N" & alteration == "Y" ~ "Alt_E",
   .default = "Unknown")) %>% 
  select(permit_type)

#make it permanent

data_2 <- data_2 %>% 
  mutate(permit_type = case_when(
   newbuilding == "Y" & electricalwork == "Y" & plumbingwork == "Y" & solar == "Y" & addition == "N" & alteration == "N" ~ "NB_EPS",
   newbuilding == "N" & electricalwork == "Y" & plumbingwork == "Y" & solar == "N" & addition == "Y" & alteration == "Y" ~ "Alt_Add_EP", 
   newbuilding == "N" & electricalwork == "Y" & plumbingwork == "Y" & solar == "N" & addition == "N" & alteration == "Y" ~ "Alt_EP",
   newbuilding == "N" & electricalwork == "Y" & plumbingwork == "N" & solar == "N" & addition == "N" & alteration == "Y" ~ "Alt_E",
   .default = "Unknown"))

```

narrow to applicants who have had 10 or more permits issued. 
```{r}
#how many rows are in the data? 
data_2 %>% count()
#there are 235934 rows in the data

data_2 %>% 
  count(planmaker) %>% 
  arrange(desc(n))
# 123682 applications have "NONE" under "planmaker" and "7077" applications have NA. That's 130,759 out of 235934. That's more than half of the applications, or 55.42%. 

#filter for only applicants with 10 or more applications
data_2 %>% 
  group_by(planmaker) %>% 
  filter(n() >= 10) %>% 
  count(planmaker) %>% 
  arrange(desc(n))

#make it permanent

data_2 <- data_2 %>% 
  group_by(planmaker) %>% 
  filter(n() >= 10)

```

Calculate the difference between each applicant's "created_date" and "issued_date." It's not clear if Civil Beat used "issued_date" or "completeddate" in their analysis. I am assuming "completeddate" is the date the actual construction was complete, not the data when the permit was issued. 

```{r}
data_2 %>% 
  mutate(days_to_issue = issue_date - created_date, .after = created_date) %>% 
  select(created_date, issue_date, days_to_issue)

#make it permanent
data_2 <- data_2 %>% 
  mutate(days_to_issue = issue_date - created_date, .after = created_date)
```

How many planmakers are in my data? 
```{r}
```{r}
data_2 %>% 
  count(planmaker)
#There are 848 unique planmakers in our data

```{r}
data_2 %>% 
  count(planmaker) %>% 
  arrange(desc(n))

#It's not clear how Civil Beat handled the NONE or NA data. I will find averages for the numbers with NONE and NA in "planamker" and averages for the data with those permit applications filtered. 
#There are 	123682	applications for which planmaker is NONE
#There are 7077 applications for which planmaker is NA
#That is 130,759 permit applications for which the planmaker is NONE or NA
```

How many permit applications (rows) are in our data?  
```{r}
data_2 %>% 
  ungroup() %>% 
  summarise(total_rows = n())

#There are 183805 total permit applications in my data. 

#How many are there without NA or NONE in the Planmaker?

data_2 %>% 
  ungroup() %>% 
  filter(is.na(planmaker) | grepl("NONE", planmaker, ignore.case=T)) %>% 
  summarise(total_rows = n())
#There are 130759 total permit applications in my data without NA and NONE
```
What is the average days to issue a permit for all our data? 

```{r}
data_2 %>% 
  ungroup() %>% 
  filter(!is.na(days_to_issue)) %>% 
  summarise(average_total_data = mean(days_to_issue))

#58.5 The average days to issue a permit was 59 days, or about two months. 
```


For each category, how many permit applications were filed? How many planmakers are there? What is the average amount of time it takes to get a permit issued? Make a list of the top 30 planmakers with the lowest average days_to_permit time. Give the average time for each, and the number of permit application each planmaker submitted. 

#1) New building with electrical, plumbing and solar (NB_EPS)

```{r}
#How many permit applications were filed? 
data_2 %>% 
  ungroup() %>% 
  filter(permit_type == "NB_EPS") %>% 
   summarise(total_rows = n())
#9,296 rows belong to NB_EPS category of permit

#How many planmakers are there? 
data_2 %>% 
  filter(permit_type == "NB_EPS") %>% 
  ungroup() %>% 
  summarise(num_planmakers = n_distinct(planmaker))
#There are 230 planmakers in this group

#What is the average amount of time it takes to get a permit of this type issued? 
data_2 %>% 
  ungroup() %>% 
  filter(!is.na(days_to_issue), permit_type == "NB_EPS") %>% 
  summarise(average = mean(days_to_issue))

#For a new building with electrical work, plumbing and solar, it took 227 days on average to issue a permit. Civil Beat's average is 265 days.

#Let's try get rid of NA and NONE 
data_2 %>% 
  ungroup() %>% 
  filter(!is.na(days_to_issue), permit_type == "NB_EPS", is.na(planmaker) | grepl("NONE", planmaker, ignore.case=T)) %>% 
  summarise(average = mean(days_to_issue))
#I now get 188.5417 days. That's even further from Civil Beat's average. 
```

Find the averages in each category 
1) New building with electrical, plumbing and solar (NB_EPS)
```{r}
#Make a list of the top 30 planmakers with the lowest average days_to_permit time. Give the average time for each, and the number of permit application each planmaker submitted. 

data_2 %>% 
  filter(!is.na(days_to_issue), permit_type == "NB_EPS") %>% 
  group_by(planmaker) %>% 
  summarise(average_days = mean(days_to_issue), applications_submitted = n()) %>% 
  select(planmaker, average_days, applications_submitted) %>% 
  arrange(desc(average_days))

#MICHAEL J KUJUBU submitted 119 permit applications and they took, on average, 1,592 days to get approved. 

data_2 %>% 
  filter(!is.na(days_to_issue), permit_type == "NB_EPS") %>% 
  group_by(planmaker) %>% 
  summarise(
    average_days = mean(days_to_issue),
    applications_submitted = n()) %>% 
  select(planmaker, average_days, applications_submitted) %>% 
  arrange(average_days) %>% 
  slice_head(n = 30)

#DARREN Y T LEE submitted two NB_EPS permits and had them approved in 4.5 days, on average. Note that I filtered earlier so the data contains only planmakers who submitted 10 or more permits. The other 8 permits DARREN Y T LEE submitted must have not fit the NB_EPS. Here's another thing I realize now that I don't know about Civil Beat's data: Did they filter for people who submitted 10 or more permits overall, or within each group of permit_type? I suspect the former because of they way they explain they data process, but this could be another reason our data results don't match.

#Let's try filter again for only applicants within the NB_EPS pool who sumbitted 10 or more NB_EPS applications. 

data_2 %>% 
  filter(!is.na(days_to_issue), permit_type == "NB_EPS") %>% 
  group_by(planmaker) %>% 
  summarise(
    average_days = mean(days_to_issue),
    applications_submitted = n()
  ) %>% 
  filter(applications_submitted >= 10) %>% 
  arrange(average_days) %>% 
  slice_head(n = 30)

#I am still not getting the same results as Civil Beat. The first applicant on their list was Makaha Oceanview Estates et al. Is that company in our list? 

data_2 %>% 
  filter(!is.na(days_to_issue), permit_type == "NB_EPS") %>% 
  group_by(planmaker) %>% 
  summarise(
    average_days = mean(days_to_issue),
    applications_submitted = n()
  ) %>% 
  filter(applications_submitted >= 10) %>% 
  arrange(planmaker)
#Makaha Oceanview Estates LLC et al. is not on our list at all. Let's get rid of NA and NONE for planmaker. 

data_2 %>% 
  filter(!is.na(days_to_issue), 
         permit_type == "NB_EPS",  
         !is.na(planmaker), 
         !grepl("NONE", planmaker, ignore.case=T)) %>% 
 group_by(planmaker) %>% 
  summarise(
    average_days = mean(days_to_issue),
    applications_submitted = n()
  ) %>% 
  filter(applications_submitted >= 10) %>% 
  arrange(average_days) 


```

2) Alteration with addition with electrical and plumbing work -- Alt_Add_EP
```{r}

data_2 %>% 
  ungroup() %>% 
  distinct(permit_type)

#How many permit applications were filed? 
data_2 %>% 
  ungroup() %>% 
  filter(permit_type == "Alt_Add_EP") %>% 
   summarise(total_rows = n())
#9,250 rows belong to Alt_Add_EP category of permit

#How many planmakers are there? 
data_2 %>% 
  filter(permit_type == "Alt_Add_EP") %>% 
  ungroup() %>% 
  summarise(num_planmakers = n_distinct(planmaker))
#There are 308 planmakers in this group

#What is the average amount of time it takes to get a permit of this type issued? 
data_2 %>% 
  ungroup() %>% 
  filter(!is.na(days_to_issue), permit_type == "Alt_Add_EP") %>% 
  summarise(average = mean(days_to_issue))

#For a new building with electrical work, plumbing and solar, it took 196 days on average to issue a permit. Civil Beat's average is 135 days.

#Get rid of NA and NONE 
data_2 %>% 
  ungroup() %>% 
  filter(!is.na(days_to_issue), permit_type == "Alt_Add_EP", is.na(planmaker) | grepl("NONE", planmaker, ignore.case=T)) %>% 
  summarise(average = mean(days_to_issue))
#The average time to issue a permit in the Alt_Add_EP permit type is 196 days. 

data_2 %>% 
  filter(!is.na(days_to_issue), permit_type == "Alt_Add_EP") %>% 
  group_by(planmaker) %>% 
  summarise(
    average_days = mean(days_to_issue),
    applications_submitted = n()) %>% 
  select(planmaker, average_days, applications_submitted) %>% 
  arrange(average_days) %>% 
  slice_head(n = 30)
```

3) Alteration with electrical and plumbing - Alt_EP
```{r}
#How many permit applications were filed? 
data_2 %>% 
  ungroup() %>% 
  filter(permit_type == "Alt_EP") %>% 
   summarise(total_rows = n())
#5,519 rows belong to Alt_EP category of permit

#How many planmakers are there? 
data_2 %>% 
  filter(permit_type == "Alt_EP") %>% 
  ungroup() %>% 
  summarise(num_planmakers = n_distinct(planmaker))
#There are 508 planmakers in this group

#What is the average amount of time it takes to get a permit of this type issued? 
data_2 %>% 
  ungroup() %>% 
  filter(!is.na(days_to_issue), permit_type == "Alt_EP") %>% 
  summarise(average = mean(days_to_issue))

#For a new building with electrical work, plumbing and solar, it took 195 days on average to issue a permit. Civil Beat's average is 135 days -- the same as it was for Alt_Add_EP 

#Get rid of NA and NONE 
data_2 %>% 
  ungroup() %>% 
  filter(!is.na(days_to_issue), permit_type == "Alt_EP", is.na(planmaker) | grepl("NONE", planmaker, ignore.case=T)) %>% 
  summarise(average = mean(days_to_issue))
#The average time to issue a permit in the Alt_Add_EP permit type is 124 days. 

data_2 %>% 
  filter(!is.na(days_to_issue), permit_type == "Alt_EP") %>% 
  group_by(planmaker) %>% 
  summarise(
    average_days = mean(days_to_issue),
    applications_submitted = n()) %>% 
  select(planmaker, average_days, applications_submitted) %>% 
  arrange(average_days) %>% 
  slice_head(n = 30)
```


4) Alteration with electrical work - Alt_E

```{r}
#How many permit applications were filed? 
data_2 %>% 
  ungroup() %>% 
  filter(permit_type == "Alt_E") %>% 
   summarise(total_rows = n())
#3,522 rows belong to Alt_E category of permit

#How many planmakers are there? 
data_2 %>% 
  filter(permit_type == "Alt_E") %>% 
  ungroup() %>% 
  summarise(num_planmakers = n_distinct(planmaker))
#There are 481 planmakers in this group

#What is the average amount of time it takes to get a permit of this type issued? 
data_2 %>% 
  ungroup() %>% 
  filter(!is.na(days_to_issue), permit_type == "Alt_E") %>% 
  summarise(average = mean(days_to_issue))

#For a new building with electrical work, plumbing and solar, it took 244 days on average to issue a permit. Civil Beat's average is 127 days

#Get rid of NA and NONE 
data_2 %>% 
  ungroup() %>% 
  filter(!is.na(days_to_issue), permit_type == "Alt_E", is.na(planmaker) | grepl("NONE", planmaker, ignore.case=T)) %>% 
  summarise(average = mean(days_to_issue))
#The average time to issue a permit in the Alt_EP permit type is 99 days. 

data_2 %>% 
  filter(!is.na(days_to_issue), permit_type == "Alt_EP") %>% 
  group_by(planmaker) %>% 
  summarise(
    average_days = mean(days_to_issue),
    applications_submitted = n()) %>% 
  select(planmaker, average_days, applications_submitted) %>% 
  arrange(average_days) %>% 
  slice_head(n = 30)
```

My attempts at recreating Civil Beat's data have thus far been unsuccessful.  