---
title: "Final Project for ADJ"
output: 
---

```{r}
library(tidyverse)
library(readxl)
library(googlesheets4)
library(janitor)
```
Loading data from the Honolulu open portal website. https://data.honolulu.gov/business/Building-Permits-January-1-2005-through-March-31-2/4vab-c87q/about_data

```{r}
read_csv("Data/Building_Permits_-_January_1__2005_through_March_31__2025_20250418 (1).csv")
original_data <- read_csv("Data/Building_Permits_-_January_1__2005_through_March_31__2025_20250418 (1).csv")
```
The data is from data.Honolulu.gov. A description from the website reads: "Building Permits - January 1, 2005 through March 31, 2025. This file is a static view of Building Permit data. This snapshot of data is an extract from the source database as of the moment of the snapshot's creation. A snapshot does not reflect changes that occur in the live data source." 

There are 428,097 rows in the data and 77 columns. I will cut down on the columns to make analyzing the data set more manageable. What columns are included?

```{r}
colnames(original_data)
```
What variables do I actually care about? 
6) contractor - who's getting permits? 
7) contractorelectrical
8) contractorplumbing
19) issuedate - are permits actually being issued? 
23) occupancygroupassessed - this tells me if it's an office building, a single-family home, an apartment, etc.
27) planmaker - this MIGHT be the person who actually submits the permit, or it could be the architect
30) proposeduse - many n/a but things like "school" would be interesting to look at
35) statusdescription - tells me if the permit is closed, the job has been canceled, or in progress. Note "sfd" is sometimes used, and "single family dwelling" at other times. Also check for trends in solar installations
36) TMK - that's the Tax Key Map, run by the city government. I can see exactly where the property is and who owns it. 
40) bpfeescollected - does this mean the permitting fees collected by the state? 
41) ownershipassessed - tells me if the property or land is private or public
42) structurecode - tells me if office building, apartment, museum, etc. 
43) date created
44) date completed
45) commercialresidential
47) ohana - that's what people in Hawaii call accessory dwelling units  
48) solar - 
52) bofeecalculation - how is this different than 40) bofeescollected? 
53) cityproject
64) stateproject
67) tmklookup - how is this different than TMK? 
76) electric vehicle charger - Check trends
77) address - last five characters are the zip code. I can create a new column of Zip code. 

Notes: There are columns for "jobaddress" and "joblocation" which often to usually line up with "address" except that many do not have zip codes at the end. 

I would like to make a new data set with only the rows I care about. I want to clean the date columns (43 and 44) so the dates appear as YYYY-MM-DD 
```{r}
original_data %>% 
  mutate(new_createddate = mdy(createddate)) %>% 
  select(new_createddate, createddate)
#this worked quite nicely. Let's try to also do the completed date. 

original_data %>% 
  mutate(new_createddate = mdy(createddate), new_completeddate = mdy(completeddate)) %>% 
  select(new_createddate, createddate, new_completeddate, completeddate)
#all of the completed dates come out at n/a. It's unclear if that's because there actually is only n/a -- I only see the first 10,000 rows -- or because something is going wrong with my code. I'll try with only completed date, then try to filter for n/as

original_data %>% 
  mutate(new_completeddate = mdy(completeddate)) %>% 
  select(new_completeddate, completeddate) %>% 
  filter(!is.na(completeddate)) %>% 
  arrange(desc(completeddate))
#There are some completeddates in 2024. That means I do want that data. For how many rows is the completeddate n.a? 

original_data %>% 
  filter(is.na(completeddate)) %>%
           count()
#there are 37,431 rows for which the completeddate is na
```
Again, I would like to make a new data set with only the rows I care about. I want to clean the date columns (43 and 44) so the dates appear as YYYY-MM-DD 

```{r}
original_data %>% 
  mutate(new_createddate = mdy(createddate), new_completeddate = mdy(completeddate)) %>% 
  select(new_createddate, createddate, new_completeddate, completeddate)

cleaned_dates_data <- original_data %>% 
   mutate(new_createddate = mdy(createddate), new_completeddate = mdy(completeddate)) %>% 
  select(new_createddate, createddate, new_completeddate, completeddate)
#I didn't want to select only those rows. 

cleaned_dates_data <- original_data %>% 
   mutate(new_createddate = mdy(createddate), new_completeddate = mdy(completeddate))

cleaned_dates_data <- original_data %>% 
   mutate(new_createddate = mdy(createddate), .after = createdate, new_completeddate = mdy(completeddate), .after = completeddate)

cleaned_dates_data <- original_data %>% 
   mutate(created_date = mdy(createddate), .after = createddate) %>% 
  mutate(completed_date = mdy(completeddate), .after = completeddate) %>% 
  mutate(issue_date = mdy(issuedate), .after = issuedate)

cleaned_dates_data %>% 
  select(buildingpermitno, contractor, contractorelectrical, contractorplumbing, issue_date, occupancygroupassessed, planmaker, proposeduse, statusdescription, tmk, bpfeescollected, ownershipassessed, structurecode, created_date, completed_date, commercialresidential, ohana, solar, bpfeecalculation, cityproject, stateproject, tmklookup, electricvehiclecharger, address)

 cleaned_data <- cleaned_dates_data %>% 
  select(buildingpermitno, contractor, contractorelectrical, contractorplumbing, issue_date, occupancygroupassessed, planmaker, proposeduse, statusdescription, tmk, bpfeescollected, ownershipassessed, structurecode, created_date, completed_date, commercialresidential, ohana, solar, bpfeecalculation, cityproject, stateproject, tmklookup, electricvehiclecharger, address, acceptedvalue, newbuilding, alteration, addition, electricalwork, plumbingwork, solar)
```

Above, I revised the cleaned_dates_data to create a data set with only the columns (variables) I care about. This will help data analysis because it will unclutter the data. I won't have to skim through variables I don't particularly care about. 

I'm interest in data that tells me: contractors in charge of the projects, dates, location(s), cost of project, cost of fees, type of property, type, electricvehiclecharger of renovation. I will clump each column into these categories. I am going to get rid of these columns: ohana and solar. I can add those back in but right now it's too much information and I don't particularly care. I am also excluding tmklookup. I am adding in: acceptedvalue

Other: buildingpermitno
Status of permit issuance: statusdescription
Contractors: planmaker, contractor, contractorelectrical, contractorplumbing
Dates: created_date, issue_date, completed_date,
Location: address, tmk
Cost of project: acceptedvalue, estimatedvalueofwork
Cost of fees: bpfeecalculation, bpfeescollected
Type of property: commercialresidential, ownershipassessed, occupancygroupassessed, proposeduse, structurecode, cityproject, stateproject
For Civil Beat analysis (see below for explanation): newbuilding, alteration, addition, electricalwork, plumbingwork, solar

```{r}
#I am further narrowing down the data I want, and the order I want to see it in. 
cleaned_data <- cleaned_dates_data %>% 
  select(buildingpermitno, statusdescription, planmaker, contractor, contractorelectrical, contractorplumbing, created_date, issue_date, completed_date, address, tmk, estimatedvalueofwork, acceptedvalue, bpfeecalculation, bpfeescollected, commercialresidential, ownershipassessed, occupancygroupassessed, proposeduse, structurecode, cityproject, stateproject)

#now I will rename columns so I can reach them more easily.
cleaned_data %>% 
  rename(building_permit_no = buildingpermitno, status = statusdescription, plan_maker = planmaker, electrical = contractorelectrical, plumbing = contractorplumbing, accepted_value = acceptedvalue, bp_fee_calculation = bpfeecalculation, bp_fee_collected = bpfeescollected, commpercial_residential = commercialresidential, ownership = ownershipassessed, occupancy_group = occupancygroupassessed, proposed_use = proposeduse, structure_code = structurecode, city = cityproject, state = stateproject)
         
cleaned_data <- cleaned_data %>% 
    rename(building_permit_no = buildingpermitno, status = statusdescription, plan_maker = planmaker, contractor_electrical = contractorelectrical, contractor_plumbing = contractorplumbing, accepted_value = acceptedvalue, bp_fee_calculation = bpfeecalculation, bp_fee_collected = bpfeescollected, commpercial_residential = commercialresidential, ownership = ownershipassessed, occupancy_group = occupancygroupassessed, proposed_use = proposeduse, structure_code = structurecode, city = cityproject, state = stateproject, est_value_of_work = estimatedvalueofwork)

#I need to add estimatedvalueofwork
```

My data is now cleaned and I am ready to analyze it. Before I move on with my cleaned data, I want to answer some questions about these four rows in original_data: are "accessorydwellingunitadu" and "ohana" ever different? Are "solar" and "solarvpinstsallation" ever different? The rows either have Y in the cell, for "yes" or N for "no." Based on skimming the data, it seems like if a row has an N for accessorydwellingunitadu that row also has an N for ohana. likewise for solarvpinstallation and solar. I want to check that that is true 100% of the time. 

Accessory dwelling units are small cottages people build in their backyard, for example, to add more living space on their property. In Hawaii, these are often called "ohana units." They're also called "granny flats."

```{r}
original_data %>% 
  select(accessorydwellingunitadu, ohana) %>% 
  filter(accessorydwellingunitadu != ohana)

#There are 3,171 rows for which "accessorydwellingunitadu" and "ohana" do not align. 
```
 
```{r}
original_data %>% 
  select(solarvpinstallation, solar) %>% 
  filter(solarvpinstallation != solar)
#there are 83,171 rows for which solarvpinstallation does not match solar. 
```

The columns "solar" and "solarvpinstallation" and not redundant. The columns "accessorydwellingunitadu" and "ohana" appear to also not be redundant. A reporting question would be how the variables within these seemingly similar pairs differ. 

The data description  says it contains data from January 1, 2005 to March 31, 2025. Is that true? 
```{r}
cleaned_data %>% 
  count(created_date)

cleaned_data %>% 
  summarise(range(created_date))
```
This contains data from 1999 to 2025, not from 2005 to 2025. 

How many rows does the data contain per year? First I'll create a new column with just the year. 

```{r}
cleaned_data %>% 
  mutate(year = year(created_date), .after = created_date)

cleaned_data <- cleaned_data %>% 
    mutate(year = year(created_date), .after = created_date)
```

The I will count the number of rows per year
```{r}
cleaned_data %>% 
  group_by(year) %>% 
  count()

#The number of permits issued in 1999 is just 364 -- hard to believe. The number rises to 1784 in 2000, then 2668 in 2001, then 4655 in 2002, 8345 in 2003, 13808 in 2003 and finally 18640 in 2005. I wonder if the data is incomplete for the years prior to 2005, seeing as the city's open-source data portal from whence I got the data say it is data from 2005 to 2025. 

cleaned_data %>%
  group_by(year) %>%
  count() %>%
  ggplot(aes(x = year, y = n)) +
geom_col()
```
I do not trust that there is complete data prior to 2005, so I will remove all the data from years prior to 2005. 

```{r}
cleaned_data %>%
  distinct(year) %>% 
  arrange(desc(year))

cleaned_data %>% 
  filter(year > 2004)

cleaned_data %>% 
  filter(year > 2004) %>% 
  distinct(year) %>% 
  arrange(desc(year))

cleaned_data <- cleaned_data %>% 
  filter(year>2004)


cleaned_data %>% 
  summarise(range(created_date))
```

The new data set, cleaned_data, has 396,473 rows and 23 columns. It shows all permit applications filed from 2005 to March 31 2025 I am ready to summarize the data. I'm going to go through and answer basic questions about the data to get a feel for what I'm working with, then I will delve into more complicated analysis questions.

How many plan makers are in the data? 

```{r}
cleaned_data %>% 
  summarise(plan_maker = n_distinct(plan_maker))

#There are just 65,987 "plan makers" for almost 400,000 permit applications filed for the Department of Planning and Permitting in Honolulu. Which plan makers appear most often? 

cleaned_data %>%
  distinct(plan_maker) 
  
cleaned_data %>%
  group_by(plan_maker) %>% 
  count() %>% 
  arrange(desc(n))

#Most plans have "NONE" for plan_maker or NA. How many of the rows have "NONE" or NA total, and what percent are they of the total? 

cleaned_data %>%
  group_by(plan_maker) %>% 
  count() %>% 
  arrange(desc(n))
  
```

I would like to reverse engineer a data analysis done by Honolulu Civil Beat as a way to practice data analysis skills. The outlet got the same data, and did the following steps, which I will replicate: 
- create civil_beat data set with only columns needed (DONE)
-narrow data to 2012 to 2022 (DONE)
- break up data into four categories for an apples-to-apples comparison of permit issuing times (DONE)
  1) New building with electrical, plumbing and solar work
  2) Alternation and addition with electrical and plumbing work 
  3) Alteration with electrical and plumbing work 
  4) Alternation with electrical work
- Narrow to applicants who have had 10 or more permits issued (DONE)
- calculate the difference between each permit application's "date created" with "date issued"  (DONE)

Following these steps, I'll create a new data set for 2023 to 2025 to see if trends have continued. 

After that, I will do further analysis to answer my own questions: 
Are solar permits increasing over time?
Which contractors are dominating city vs. state jobs?
Do residential permits get issued faster than commercial ones?

Replicating Civil Beat's Analysis: 
1) add columns to "cleaned_data": "newbuilding," "electricalwork", "solar", "addition", "alternation", "plumbingwork"
```{r}
colnames(cleaned_data)
#1) I need to add: "newbuilding," "electricalwork", "solar", "addition", "alternation", "plumbingwork" - those columns exist in original_data, but not in clean_data. rather than doing a left_join, I'll just go back in my code where I selected the columns I wanted from cleaned_dates_data to add into cleaned_data. 

ncol(cleaned_data)
#I went back up and added the columns, and now have 30 columns in my data. 
colnames(cleaned_data)
```

I will keep "cleaned_data" for later analysis and create a new data set called "civil_beat" to create a data set with only the columns I need to re-create their analysis. Here are the variables (columns) I'll need: 
"buildingpermitno" , “contractor" "issue_date"  , "occupancygroupassessed" , "planmaker"              "proposeduse" ,”statusdescription”, "bpfeescollected" "created_date" , "commercialresidential" ,”solar", "bpfeecalculation" ,”address" ,”acceptedvalue”,”newbuilding”,”alteration”,”addition","electricalwork" , "plumbingwork"
```{r}
civil_beat <- cleaned_data %>% 
  select(buildingpermitno, contractor, issue_date, occupancygroupassessed, planmaker, proposeduse, statusdescription, bpfeescollected, created_date, commercialresidential, solar, bpfeecalculation, address, acceptedvalue, newbuilding, alteration, addition, electricalwork, plumbingwork)

colnames(civil_beat)
```

I'll narrow the data down to 2012 to 2022
```{r}
civil_beat %>% 
  mutate(year = year(created_date), .after = created_date) %>% 
  select(year, created_date)

civil_beat<-civil_beat %>% 
  mutate(year = year(created_date), .after = created_date)

civil_beat %>% 
  filter(year > 2011) %>% 
  range(year)

civil_beat %>% 
  filter(year > 2011) %>% 
  distinct(year) %>% 
  arrange(desc(year))

civil_beat %>% 
  filter(year > 2011 & year < 2023) %>% 
  distinct(year) %>% 
  arrange(desc(year))

civil_beat %>% 
  filter(year > 2011 & year < 2023) %>% 
  count(year) %>% 
  arrange(desc(year))

range(civil_beat$created_date)

civil_beat <- civil_beat %>% 
  filter(year > 2011 & year < 2023)
  
```

break up data into four categories for an apples-to-apples comparison of permit issuing times
  1) New building with electrical, plumbing and solar work
  2) Alternation and addition with electrical and plumbing work 
  3) Alteration with electrical and plumbing work 
  4) Alternation with electrical work
  
```{r}
civil_beat %>% 
  mutate(permit_category = case_when(
         newbuilding == "Y" & electricalwork == "Y" & plumbingwork == "Y" & solar == "Y"~ "NB_EPS",
         alteration=="Y"& addition=="Y" & electricalwork=="Y" & plumbingwork == "Y" ~ "Alt_Add_EP", 
         alteration=="Y" & plumbingwork=="Y" & electricalwork == "Y" ~ "Alt_EP",
         alteration == "Y" & electricalwork == "Y" ~ "Alt_E", 
         TRUE ~ NA_character_))

civil_beat <- civil_beat %>%
    mutate(permit_category = case_when(
         newbuilding == "Y" & electricalwork == "Y" & plumbingwork == "Y" & solar == "Y"~ "NB_EPS",
         alteration=="Y"& addition=="Y" & electricalwork=="Y" & plumbingwork == "Y" ~ "Alt_Add_EP", 
         alteration=="Y" & plumbingwork=="Y" & electricalwork == "Y" ~ "Alt_EP",
         alteration == "Y" & electricalwork == "Y" ~ "Alt_E", 
         TRUE ~ NA_character_))
```

The data now contain 235,934 rows. I don't want to look at projects that do not fit into my criteria, in other words, projects with "na" in my permit_category column. 
```{r}
civil_beat %>% 
  filter(!is.na(permit_category))
#that brought my number of rows down to 47,935. 

civil_beat <- civil_beat %>% 
  filter(!is.na(permit_category))
```

Now I need to narrow to applicants who have had 10 or more permits issued. Should I use "contractor" or "planmaker" for applicant? Let's compare these two rows. 

```{r}
civil_beat %>% 
  select(planmaker, contractor)
#There are nas in both columns, and a number of "nones" in the contractor column. How many nas are in the planner column and how many are in the contractor column? 

civil_beat %>% 
  filter(is.na(planmaker))
#734 rows in the data have na for planmaker

civil_beat %>% 
  filter(is.na(contractor))
#5,372 rows have na for contractor column. 
```

Now I need to narrow to applicants who have had 10 or more permits issued. I will use "planmaker" for applicant. I need to group_by planmaker and then filter for planmakers who have less than 10 rows. 
```{r}
civil_beat %>% 
  group_by(planmaker) %>% 
  arrange(desc(planmaker)) %>% 
  select(planmaker)

civil_beat %>% 
  count(planmaker) %>% 
  arrange(desc(n))

 civil_beat %>% 
   group_by(planmaker) %>% 
   filter(n() >= 10)
#I am now working with 28,807 rows of data
 
 civil_beat %>% 
   group_by(planmaker) %>% 
   filter(n() >= 10) %>% 
   summarise(num=n())

civil_beat %>% 
   group_by(planmaker) %>% 
   filter(n() >= 10) %>% 
   summarise(planmaker = n_distinct(planmaker)
             
civil_beat %>% 
  group_by(planmaker) %>% 
  filter(n() >= 10) %>% 
  ungroup() %>% 
  summarise(num_planmakers = n_distinct(planmaker))
 
#there are 410 planmakers in our data -- people whose plans match our criteria and who have submitted more than 10 applications between 2012 and 2022. 

civil_beat <- civil_beat %>% 
  group_by(planmaker) %>% 
  filter(n() >= 10)

```

My next step: calculate the difference between each permit application's "date created" with "date issued" . Here are the steps: create a new column (mutate) and within that column put the calculated number of days between "date created" and "date issued." In the same code chunk, after mutate, add a formula to subtract created_date from issued_date. This will be my first time calculating the number of days between two dates in R. 

```{r}
civil_beat %>% 
  mutate(date_to_issue = issue_date - created_date)

civil_beat <- civil_beat %>% 
  mutate(date_to_issue = issue_date - created_date, .after = issue_date)

civil_beat %>% 
  filter(date_to_issue < 0)
#no rows have negative values

civil_beat %>% 
  filter(date_to_issue == 0)

# 43 permit applications were approved on the same day between 2012 and 2022. 

#just out of curiosity, who got those permits? 
civil_beat %>% 
  filter(date_to_issue == 0) %>% 
  count(planmaker) %>% 
  arrange(desc(n))
#"NONE" got 17 permits issued in a day. FISHY. Methode Maniraguha got 7 permits issued in just one day, and infamous Honolulu monster home builder Xiang Yee had 3 permits issued the same day

civil_beat %>% 
  filter(date_to_issue == 0) %>% 
  count(planmaker, permit_category) %>% 
  arrange(desc(n))
```

Find the average and median days to issue for each planmaker across all permit types from the categories you created. 
```{r}
civil_beat %>% 
  summarise(median_days_to_issue = median(date_to_issue, na.rm = TRUE))%>% 
  arrange((median_days_to_issue))

group_vars(civil_beat)

civil_beat %>% 
  summarise(mean_days_to_issue = mean(date_to_issue, na.rm = TRUE))%>% 
  arrange(mean_days_to_issue)

```


Find the average and median days to issue across all permit types from the categories you created. 
```{r}
civil_beat %>% 
  ungroup() %>% 
  summarise(median_days_to_issue = median(date_to_issue, na.rm = TRUE))

#Median days to issue is 125 days across all project types
group_vars(civil_beat)

civil_beat %>% 
    ungroup() %>% 
  summarise(mean_days_to_issue = mean(date_to_issue, na.rm = TRUE))
#average (mean) days to issue is 210 days	across all project types 
```
Next: find mean and median for each permit_category. (filter for each permit category and find the mean and median)
Check outliers to see people who have gotten their permits approved in far shorter time. 
Permit categories: 
  1) NB_EPS = New building with electrical, plumbing and solar work
  2) Alt_Add_EP = Alternation and addition with electrical and plumbing work 
  3) Alt_EP = Alteration with electrical and plumbing work 
  4) Alt_E = Alternation with electrical work
  
  What is the mean and median time to issue a permit for NB_EPS permits? 
```{r}
civil_beat %>% 
  group_by(permit_category) %>% 
  summarise(median_time = median(date_to_issue))

civil_beat %>% 
  filter(!is.na(date_to_issue)) %>% 
  group_by(permit_category) %>% 
  summarise(median_time = median(date_to_issue))

#Civil Beat's analysis used averages, not mean 

civil_beat %>% 
  filter(!is.na(date_to_issue)) %>% 
  group_by(permit_category) %>% 
  summarise(average = mean(date_to_issue))
```
For Alt_Add_EP, Civil Beat's average is 135 days. My average is 199 days. 
For Alt_E, Civil Beat's average is 127 days. My average is 220 days.
For Alt_EP, Civil Beat's average is 135 days. My average is 193 days.
For NB_EPS, Civil Beat's average is 265 days. My average is 227 days.

I'm going to go back and look through my code for potential errors. What steps did I take? 
- create civil_beat data set with only columns needed
        Checked: I added the necessary columns to create civil_beat
-narrow data to 2012 to 2022 
     Checked: I used range(civil_beat$created_date) to figure out the dates. I got: [1] "2012-01-03" "2022-12-31"
      
- break up data into four categories for an apples-to-apples comparison of permit issuing times
  1) New building with electrical, plumbing and solar work
  2) Alternation and addition with electrical and plumbing work 
  3) Alteration with electrical and plumbing work 
  4) Alternation with electrical work
Here's the code I used to created the four categories: 
civil_beat %>% 
  mutate(permit_category = case_when(
         newbuilding == "Y" & electricalwork == "Y" & plumbingwork == "Y" & solar == "Y"~ "NB_EPS",
         alteration=="Y"& addition=="Y" & electricalwork=="Y" & plumbingwork == "Y" ~ "Alt_Add_EP", 
         alteration=="Y" & plumbingwork=="Y" & electricalwork == "Y" ~ "Alt_EP",
         alteration == "Y" & electricalwork == "Y" ~ "Alt_E", 
         TRUE ~ NA_character_))
Here's what I think could have gone wrong. In creating the group "Alteration with electrical work," for example, I didn't put in provisions to say the group should have "N" for thinkgs like Solar, or Addition, or Plumbing. So is it possible that some of the permit applications in my "Alternation with electrical work" are ALSO projects that have plumbing work becuase I didn't specify that those projects should NOT have plumbing? 
      CHECKED: This was, indeed, the problem. 

One way to check this is to see: How many permits are in each group and how many are there total in all four groups. If the sum of the parts is more than the whole, I have some overlap. 

```{r}
civil_beat %>% 
  group_by(permit_category) %>%
    summarise(total_permit = n())

civil_beat %>% 
  group_by(permit_category) %>%
    summarise(total_permit = n()) %>% 
  mutate(total = sum(total_permit))

#Oops. Another lapse in my thinking. Of course the total is 28,807 -- I'm just adding up the totals. Here's what I've learned: (case_when) only allows one permit_category per row. According to ChatGPT: "R assigns the category from the first condition that matches, and then stops checking the others." Here's my confusion: So does that mean that if there's a permit that has alternation with electrical work, it could have ended up in "alteration with electrcial and plumbing?" If that's case, why are there ANY rows accounted for in my alteration with electrical work category? Wouldn't all those have been accounted for in my alteration with electrical work and plumbing category, which preceded the alternation with electrical work in my code? 

#permit_category total_permit  total
#Alt_Add_EP	      9458	     28807		
#Alt_E	          4749	    28807		
#Alt_EP	          5323    	28807		
#NB_EPS	          9277	    28807	

```
Check for NB_EPS 
```{r}
civil_beat %>% 
  filter(newbuilding == "Y", electricalwork == "Y", plumbingwork == "Y", solar == "Y") %>% 
  nrow()
#I get 9,277

civil_beat %>%
  filter(newbuilding == "Y", electricalwork == "Y", plumbingwork == "Y", solar == "Y",
         alteration == "N", addition == "N") %>%
  nrow()
#I get 9231

#46  rows were new buildings with solar, electrical and plumbing and also either alteration or addition. We don't know if Civil Beat excluded alterations and additions in their categorizing of new buildings with solar, plumbing and electrical work. 
```
Alt_Add_EP. I got 9458 earlier. 
```{r}
civil_beat %>%
  filter(alteration == "Y", addition == "Y", electricalwork == "Y", plumbingwork == "Y") %>%
  nrow()
#the answer I get is 9470

civil_beat %>%
  filter(
    alteration == "Y",
    addition == "Y",
    electricalwork == "Y",
    plumbingwork == "Y",
    newbuilding == "N",
    solar == "N"
  ) %>%
  nrow()

#the answer I get is 9101
```

```{r}
civil_beat <- civil_beat %>% 
  mutate(permit_category = case_when(
         newbuilding == "Y" & electricalwork == "Y" & plumbingwork == "Y" & solar == "Y" & alteration == "N" & addition == "N"~ "NB_EPS",
         alteration=="Y"& addition=="Y" & electricalwork=="Y" & plumbingwork == "Y" & solar == "N" & newbuilding == "N" ~ "Alt_Add_EP", 
         alteration=="Y" & plumbingwork=="Y" & electricalwork == "Y" & addition == "N" & solar == "N" & newbuilding == "N" ~ "Alt_EP",
         alteration == "Y" & electricalwork == "Y" & plumbingwork == "N" & addition == "N" & solar == "N" & newbuilding == "N" ~ "Alt_E", 
         TRUE ~ NA_character_))
```

Next I will check that I: 
- Narrow to applicants who have had 10 or more permits issued 
```{r}
 civil_beat %>% 
   group_by(planmaker) %>% 
   filter(n() < 10)
#no planmakers in my data submitted less than 10 permits

 civil_beat %>% 
   group_by(planmaker) %>% 
   filter(n() == 10)
 #34 planmakers submitted 10 applications
 
 civil_beat %>% 
   group_by(planmaker) %>% 
   filter(n() > 10)
 #376 planmakers submitted more than 10 -- that makes my 410 total plan makers who submitted 10 or more permits. 
 
```

Calculate the difference between each permit application's "date created" with "date issued". Here are some places this could have gone wrong: did Civil Beat include the date when the permit was created and issued, meaning is it inclusive? My guess is it was. 
```{r}
civil_beat %>% 
  ungroup() %>% 
  select(buildingpermitno, created_date , issue_date, date_to_issue)

civil_beat %>% 
  select(buildingpermitno, created_date , issue_date, date_to_issue)
```

Then re-calculate the mean for the 2012 - 2022 time frame. 
```{r}
civil_beat %>% 
  group_by(permit_category) %>% 
  summarise(mean = mean(date_to_issue, na.rm = TRUE))

civil_beat %>% 
  filter(!is.na(permit_category)) %>% 
  group_by(permit_category) %>% 
  summarise(average = mean(date_to_issue, na.rm = TRUE))
```
For Alt_Add_EP, Civil Beat's average is 135 days. My average was 199 days. My new average is 195 days. 
For Alt_E, Civil Beat's average is 127 days. My average was 220 days. My new average is 240 days. 
For Alt_EP, Civil Beat's average is 135 days. My average is 193 days. My new average is 192 days. 
For NB_EPS, Civil Beat's average is 265 days. My average is 227 days. My new average is 227 days. 

Ok, my hypothesis that not adding things like plumbingwork == "N" for the Alt_E category, for example, did not appear to be correct. Even with those new category stipulations when doing case_when, I still did not get what Civil Beat got, and in fact doing that did not change my averages very much. 



My own interest: Annual average and median days to usse
***as a rule of thumb, use `group_by()` and `summarise()` together, or `count()` if what you want to summarise is just number of rows**
data_2 %>% 
  group_by(ExtNatureDisplayName) %>% 
  count() %>% 
  arrange(desc(n))
###LIZ: to tighten your code a bit:
data_2 %>% 
  count(ExtNatureDisplayName) %>% 
  arrange(desc(n))

Possible Questions: 
Median time for permitting residential and commerical projects? 
What's the best day of the week to file a permit? 
For every year, what's the median time it takes to issue a Honolulu permit? 
Who gets the permits? 
Is estimated value of work (row 12) always the same as accepted value (row 1) ?
Which plumbing , electrical and general contractors are working on state and city projects most often? 
How many electric vehicle chargers have been added to homes? Check the variable: electricvehiclecharger 